Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-11-14 Paper/ms.tex
Spikes and slabs

\subsection{Inferring the number of planets}
The number of orbiting planets, $N$, is an important parameter.
To calculate the posterior distribution for $N$, most authors
consider various trial values of $N$, and calculate the marginal likelihood
\begin{eqnarray}
p(\bdata | N) &=& \int p(\btheta | N) p(\bdata | \btheta, N) \, d^n \btheta
\end{eqnarray}
for each possible value of $N$
\citep[e.g.][]{2011MNRAS.415.2523G, 2011MNRAS.415.3462F, 2014MNRAS.437.3540F, fengji}, marginalising
over all other model parameters.
The posterior distribution for $N$ can then be found straightforwardly by
using Bayes' rule with $N$ as the only unknown parameter:
\begin{eqnarray}
p(N | \bdata) &=& \frac{p(N)p(\bdata | N)}{\sum_N p(N)p(\bdata | N)}.
\end{eqnarray}
Popular
methods for calculating the marginal likelihood are Nested Sampling
\citep{skilling} and ideas related to thermodynamic integration
\citep[e.g.][]{neal}. Note that thermodynamic integration can easily fail if
there is a phase transition in the problem \citep{skilling}.

@@

To fit a planet model to RV data, we need parameters to describe
the properties of each planet. For simplicity, we describe each planet by
five parameters: $T_i$, the orbital period, $A_i$, the amplitude (in metres
per second) of the RV signal, $\phi_i$, the phase of the signal, $v_i$, an
``initial velocity'' parameter describing the eccentricity of the orbit
(see Section~\ref{sec:orbits}), and $\theta_i$, a ``viewing angle'' parameter
describing the position of the observer with respect to the major axis of the
star's elliptical trajectory.

>>>>

\subsection{Inferring the number of planets}
The number of orbiting planets, $N$, is an important parameter.
To calculate the posterior distribution for $N$, most authors
consider various trial values of $N$, and calculate the marginal likelihood
\begin{eqnarray}
p(\bdata | N) &=& \int p(\btheta | N) p(\bdata | \btheta, N) \, d^n \btheta
\end{eqnarray}
for each possible value of $N$
\citep[e.g.][]{2011MNRAS.415.2523G, 2011MNRAS.415.3462F, 2014MNRAS.437.3540F, fengji}, marginalising
over all other model parameters.
The posterior distribution for $N$ can then be found straightforwardly by
using Bayes' rule with $N$ as the only unknown parameter:
\begin{eqnarray}
p(N | \bdata) &=& \frac{p(N)p(\bdata | N)}{\sum_N p(N)p(\bdata | N)}.
\end{eqnarray}
Popular
methods for calculating the marginal likelihood are Nested Sampling
\citep{skilling} and ideas related to thermodynamic integration
\citep[e.g.][]{neal}.

@@

\subsection{Phase transitions}
It is well known that Bayesian computation (using MCMC for example) can be
difficult when the posterior distribution is multimodal or has strong
dependencies between parameters. An uncommon but less well-known difficulty
is the existence of {\it phase transitions} \citep{skilling}.

Imagine a high-dimensional unimodal posterior distribution that is composed of a
broad, high volume but low density ``slab'' with a narrow, low volume but high
density ``spike'' on top of it.
An example is a mixture of two concentric high-dimensional gaussians with
different widths. If you ran MCMC on such a posterior, it would be difficult
to jump between the slab and the spike components. If the MCMC is currently
in the spike region (or phase) it will be unable to escape: a proposed move
into the slab will be rejected because of the ratio of densities. Conversely,
if the MCMC was in the slab region, it would be unlikely to go into the spike
region, because its volume is so small: it would be very unlikely to
{\it propose} to move into the spike.

\subsection{Parameters and Priors}
To fit a planet model to RV data, we need parameters to describe
the properties of each planet. For simplicity, we describe each planet by
five parameters: $T_i$, the orbital period, $A_i$, the amplitude (in metres
per second) of the RV signal, $\phi_i$, the phase of the signal, $v_i$, an
``initial velocity'' parameter describing the eccentricity of the orbit
(see Section~\ref{sec:orbits}), and $\theta_i$, a ``viewing angle'' parameter
describing the position of the observer with respect to the major axis of the
star's elliptical trajectory.
